---
title: 'dplr: Practice'
author: "R Workshop"
output:
  html_document:
    css: ../workshop.css
    highlight: pygments
    theme: cerulean
  pdf_document: default
---

```{r, echo=FALSE, warning=FALSE, message=FALSE}
require(knitr)
require(dplyr)
opts_chunk$set(warning=FALSE, message=FALSE)
```


### Earthquake data

```{r load-quakes, eval = TRUE}
data(quakes)
```

1.   Show a command you could use to find the average (mean) magnitude of all earthquakes that occurred below the median depth of all earthquakes.

```{r}
quakes %>%
  dplyr::filter(depth < median(quakes$depth)) %>%
  summarize(mean(mag))
```

1.   How does the average magnitude of these "deep" earthquakes compare to the magnitude of "shallow" (above median depth) earthquakes?

```{r}
quakes %>%
  filter(depth > median(depth)) %>%
  summarize(mean(mag))
```

1.   Show the command you would use to get the row(s) with the smallest value in the `depth` column.

```{r}
quakes %>%
  filter(depth == min(depth))
```

1.   Show the command you would use to create a new data frame named `quakes_sorted` that stores the data in sorted order by `depth`.

```{r}
quakes_sorted <- arrange(quakes, depth)
```

* * *

### Flights data

```{r load-flights, eval = TRUE}
library(pnwflights14)
data(flights)
```

1. Say I fly very often between PDX airport in Portland and Logan Airport in Boston. Based on the 2013 data, which day of the week would you recommend that I buy a ticket for if I want to minimize delays?

```{r}
get_day_of_week <- function (d, m, y) {
  weekdays(as.Date(paste(d, "-", m, "-", y, sep = ""),
                   '%d-%m-%Y'))
}
get_day_of_week(16, 8, 2012)

a <- flights %>%
  filter((dest == 'PDX' | dest == 'BOS') & (origin == 'PDX' | origin == 'BOS')) %>%
  mutate(day_of_week = get_day_of_week(day, month, year)) %>%
  group_by(day_of_week) %>%
  summarise(count = n(), avg_delay = mean(dep_delay, na.rm = TRUE)) %>%
  arrange(avg_delay)
```

1. What advice do you have for the best airline for this route?

```{r}
flights %>%
  filter((dest == 'PDX' | dest == 'BOS') & (origin == 'PDX' | origin == 'BOS')) %>%
  group_by(carrier) %>%
  summarize(count = n(), avg_delay = mean(dep_delay, na.rm = TRUE)) %>%
  arrange(avg_delay)

```


* * *

### Back to Global Warming

Go back to the vector `tData` that you used for the project. Use this data and
maybe `dplyr` to form a dataframe that will make the analysis much more straightforward.

```{r}
load(url('http://www.stolaf.edu/people/olaf/cs125/project1.RData'))

tData_df <- data.frame(month = rep(1:12),
                       year = rep(1881:2012, each = 12),
                       tData)
```
