---
title: "Dataframes: Practice"
author: "R Workshop"
output:
  html_document:
    css: ../workshop.css
    highlight: pygments
    theme: cerulean
  pdf_document: default
---

```{r, echo=FALSE, warning=FALSE, message=FALSE}
require(knitr)
opts_chunk$set(warning = FALSE, message = FALSE)
```

For the following exercises, load in the built in earthquake data set.

```{r, eval=FALSE}
data(quakes)
```

1.   How would you find information about this data frame? How many rows and columns does it have?
```{r}
str(quakes)
```
1.   Based purely on the description of the columns (found with `?quakes`), would you expect a relationship between the 4th and 5th columns? Why or why not?

Yes. The more stations that detect the quake, the larger the earthquake's magnitude.

1.   Show a command you could use to find the average (mean) magnitude of all earthquakes that occurred below the median depth of all earthquakes.

```{r}
mean(quakes[quakes$depth < median(quakes$depth), ]$mag)
```

1.   How does the average magnitude of these "deep" earthquakes compare to the magnitude of "shallow" (above median depth) earthquakes?

```{r}
mean(quakes[quakes$depth > median(quakes$depth), ]$mag)
```

The mean recorded magnitude is lower.

1.   Show the command you would use to get the row(s) with the smallest value in the `depth` column.

```{r}
quakes[which.min(quakes$depth), ]
```

<div id="license">
Adapted from material developed by Olaf Hall-Holt and Paul Roback for CS 125 at St. Olaf College.
</div>