---
title: "Lubridate"
author: "Eddie"
date: "June 8, 2016"
output: ioslides_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)

require(lubridate)

```

# Lubridate

A great package, with an unappealing name!

## The Problem

Dates aren't standard at all!

- January 20th, 2016
- 1/20/2016
- 20-01-2016
- 20160120

## The Solution

```{r echo=TRUE}
mdy("1/20/2016")
dmy("20-01-2016")
ymd("20160120")
mode(mdy("1/20/2016"))
```

## Deconstructing Lubridates

```{r, echo=TRUE}
t <- now()
t

year(t)
month(t)
```

## Deconstructing Lubridates (cont)

```{r, echo=TRUE}
day(t)
hour(t)
minute(t)
```

## Deconstructing Lubridates (cont)

```{r, echo=TRUE}
second(t)
tz(t)

t <- now("GMT")
tz(t)
```

## Other date/time info

Get day of the week (1 = Sunday, 2 = Monday, etc)
```{r, echo=TRUE}
t <-  now()
wday(t)
```

Get day of the year
```{r, echo=TRUE}
yday(t)
```

## Change Time Zone

Convert time to different time zone
```{r, echo=TRUE}
with_tz(t, tzone = "GMT")
```

Change time zone
```{r, echo=TRUE}
force_tz(t, tzone = "GMT")
```

## Durations

How long have we been working for MassMutual
```{r, echo=TRUE}
t1 <- ymd_hms("2016-06-01-09-00-00", tz="EST")
t2 <- now()
t2 - t1
```

When will we have worked for MassMutual for 100 days?
```{r, echo=TRUE}
t1 + days(100)
```

## Durations (cont)

```{r, echo=TRUE}
t1 + milliseconds(100)
t1 + seconds(100)
t1 + minutes(100)
```
