---
title: "Project: vectors and functions"
author: "R Workshop"
output:
  html_document:
    css: ../workshop.css
    highlight: pygments
    theme: cerulean
  pdf_document: default
---

```{r, echo=FALSE, warning=FALSE, message=FALSE}
require(knitr)
opts_chunk$set(warning=FALSE,message=FALSE)
```

The following link will load the data:

```{r, echo=FALSE}
load(url('http://www.stolaf.edu/people/olaf/cs125/project1.RData'))

```

#### Recall:

- The first element of the vector corresponds to the average temperature 
in January of 1881. The second element is February 1881, etc. The last element is December 2012.
- Each element in the vector is the scaled deviation (by 100) from a baseline mean 
global temperature in Celsius. The baseline mean global temperature is the 
global temperature from 1951-1980, which is 14.0o C (57.2o Fahrenheit).
- For data details, read the bottom of the Nasa page:
http://data.giss.nasa.gov/gistemp/tabledata_v3/GLB.Ts+dSST.txt

#### Some requests

- If there are some tasks that you can imagine needing to code up again, *write a 
function for it*.
- Your final html document should read like an informative analysis of this
temperature data set meant to be consumed by a non-programmer. That is,
    - Supress R code (though an exception may be easy-to-read output).
    - The narrative should be as in a natural analysis; not as a collection of
    enumerated answers to exercises.
    - Use sensible Markdown sytax for organization.
    - Fiddle with "knitr chunk options" to get the figures you're going for.
- Your source, however, should be crystal clear to another programmer.
    + Code with style.
    + Label your chunks.

* * *

Please complete the following exercises using R vectors:

1.   Convert the data from scaled deviation in Celsius to absolute temperature in 
Fahrenheit. What is the mean temperature in Fahrenheit over the entire data set? 
Confirm that your result is reasonable. Use your new Fahrenheit vector for the 
rest of the exercises.


```{r mean_temperature, echo=FALSE}
temp_data <- tData

# Converting to F
temp_data_f <- (temp_data/100)*1.8 + 57.2

# Calculate mean
mean_temp <- mean (temp_data_f)

mean_temp

```

-Global Temperatures -
<br>
This project studies the average global temperatures starting from January 1881 and ending in December 2012. The data was parsed and analyzed to get some insights on average global temperatures. The mean temperature of the entire data set was found to be `r signif(mean_temp, digits=4)`
 

1.   Find the mean global temperature of each the summer months: June, July, and August. Also please find the lowest and highest global average temperatures for each month. This is easiest if you create new vectors that have all June temperatures, all July temperatures, and so on. 
When did the lowest and highest occur for each month? Are the highest temperatures 
generally more recent than the lowest temperatures?


```{r mean_summers, echo=FALSE}
temp_matrix <- matrix(temp_data_f, nrow = 12, ncol = 132, byrow = F)
rownames(temp_matrix) <- c('Jan', 'Feb', 'Mar', 'Apr', 'May', 'Jun', 'Jul', 'Aug', 'Sep', 'Oct', 'Nov', 'Dec')
colnames(temp_matrix) <- 1:132

# Mean of each summer
summer <- temp_matrix[6:8,]
summer_means <- rowMeans(summer)
summer_means

# Max for each month
max_temps_per_month <- apply(temp_matrix, 1, max)
max_temps_per_month

# Sort max-temp from low-high
hotest_months<-sort(max_temps_per_month)

# Min for each month
min_temps_per_month <- apply(temp_matrix, 1, min)
min_temps_per_month

# Sort min-temp from low-high
coldest_months<-sort(min_temps_per_month)

# Which years were hottest
hottest_years <- apply(temp_matrix, 1, which.max) + 1881
hottest_years

# Min for each summer
coldest_years <- apply(temp_matrix, 1, which.min) + 1881
coldest_years
```

The mean global temperature for June, July, and August (summer months) was calculated to be `r signif(summer_means[[1]], digits = 4)`, `r signif(summer_means[[2]], digits = 4)`, and `r signif(summer_means[[3]], digits = 4)` respectively. The highest temperatures for all months ranged from `r signif(hotest_months[[1]], digits = 4)` and `r signif(hotest_months[[12]], digits = 4)`. The lowest happening in July 2012 while the highest one occurred in January 2008. On the other hand, the lowest temperatures ranged from `r signif(coldest_months[[1]], digits = 4)` to `r signif(coldest_months[[12]], digits = 4)` which happened in January 1894 and May 1918 respectively, which shows that highest temperatures are more recent than colder ones. 

The table below shows the highest temperatures in Farenheit per month and the year they happened.
`r signif(rbind(max_temps_per_month, hottest_years), digits = 4)`

The table below shows the lowest temperatures in Farenheit per month and their corresponding year.
`r signif(rbind(min_temps_per_month, coldest_years), digits = 4)`
--- end ----


1.   For the rest of the project, it will be helpful to define two functions. 
Define one function that takes the month and year as input and outputs the index in 
the vector for that given month and year. Define another function that takes the index 
in the vector as the input and outputs the month and year.

``` {r utility_functions, echco=FALSE}
month_year_to_index <- function(month = 1, year = 1881) {
  y <- year - 1881
  return(y*12 + month)
}

index_to_month_year <- function(ind = 1) {
  month <- (ind %% 12)
  y <- floor(ind / 12)
  return(c(month, y+1881))
}

```

Some examples to demonstrate the functions:


```{r} 
month_year_to_index(1, 1881)
month_year_to_index(2, 1881)
month_year_to_index(1, 1882)

index_to_month_year(1)
index_to_month_year(2)
index_to_month_year(13)

```


1.   Create plots of the monthly average temperature over time for each month: 
one plot for all January average temperatures, one plot for all February average 
temperatures, etc. To make a plot with a single vector, use the `plot()` function.
The plots should have *Years* on the horizontal axis, and *Temperature* on the vertical 
axis. Does there seem to be a warming trend? The previously defined functions will 
be helpful here.  Feel free to fiddle with the optional arguments to `plot()` (more
are found with `?par`).

--- Added 

The plots below show the montHly average temperature from 1881 to 2012 time period. The plots have years in the horizontal axis and temperatures in the vertical axis. As we can see, the plots show an increasing trend. 
----- End

```{r plot_average_temp, echo=FALSE}

  years <- c(1881:2012)
  plot(years, temp_matrix["Jan",], ylab = "Temperature (F)", xlab = "Years", pch = 19, col = "blue", cex = 0.3, main= "January Average Temperatures")
  
  
  plot(years, temp_matrix["Feb",], ylab = "Temperature (F)", xlab = "Years", pch = 19, col = "dark red", cex = 0.3, main= "February Average Temperatures")
  
  plot(years, temp_matrix["Mar",], ylab = "Temperature (F)", xlab = "Years", pch = 19, col = "dark green", cex = 0.3, main= "March Average Temperatures")
  
  plot(years, temp_matrix["Apr",], ylab = "Temperature (F)", xlab = "Years", pch = 19, col = "dark blue", cex = 0.3, main= "April Average Temperatures")
  
  plot(years, temp_matrix["May",], ylab = "Temperature (F)", xlab = "Years", pch = 19, col = "gold", cex = 0.3, main= "May Average Temperatures")
  
  plot(years, temp_matrix["Jun",], ylab = "Temperature (F)", xlab = "Years", pch = 19, col = "brown", cex = 0.3, main= "June Average Temperatures")
  
  plot(years, temp_matrix["Jul",], ylab = "Temperature (F)", xlab = "Years", pch = 19, col = "purple", cex = 0.3, main= "July Average Temperatures")
  
  plot(years, temp_matrix["Aug",], ylab = "Temperature (F)", xlab = "Years", pch = 19,  cex = 0.3, main= "August Average Temperatures")
  
  plot(years, temp_matrix["Sep",], ylab = "Temperature (F)", xlab = "Years", pch = 19, col = "dark grey", cex = 0.3, main= "September Average Temperatures")
  
  plot(years, temp_matrix["Oct",], ylab = "Temperature (F)", xlab = "Years", pch = 19, col = " orange", cex = 0.3, main= "October Average Temperatures")
  
  plot(years, temp_matrix["Nov",], ylab = "Temperature (F)", xlab = "Years", pch = 19, col = " magenta", cex = 0.3, main= "November Average Temperatures")
  
  plot(years, temp_matrix["Dec",], ylab = "Temperature (F)", xlab = "Years", pch = 19, col = "brown", cex = 0.3, main= "December Average Temperatures")
```

1.   Create a plot that shows the average annual temperature over all years. 
(That is, create one plot where each data point corresponds to the average temperature 
for a year.) In what years did the highest and lowest average temperatures occur?

```{r annual_temp, echo=FALSE}
plot(years, colMeans(temp_matrix), ylab = "Temperature (F)", main= "Average Temperature (1881-2012)", xlab = "Years", pch = 19, col = "brown", cex = 0.3)

# Calculate which year had the max temperature
hottest_years<- which.max(colMeans(temp_matrix)) + 1881

# Calculate which year had the min temperature
coldest_years <- which.min(colMeans(temp_matrix)) + 1881

```

--- Added
The plot above shows the average annual temperature over all years. There is an increase in temperature trend starting from 1920. In addition there is a constant increase starting in 1960. 
The year with highest temperature was `r hottest_years[[1]]`. The year with lowest temperature was `r coldest_years[[1]]`.
-----End


1.   Create a plot showing the monthly average temperature starting at January 
2000 (one plot showing all data points from January 2000 through December 2012). 
Do you see any pattern to the data? Are temperatures rising? The previously defined 
functions will be helpful here.

```{r average_temp, echo=FALSE}

  temp_m_2k<- temp_matrix[, 120:132]

  plot(1:156, as.vector(temp_m_2k), ylab = "Temperature (F)", xlab ="Months (Jan 2000- Dec 2012)", pch = 19, col = "brown", cex = 0.3, main = "Monthly Average Temperature")

  
```

---- Added
The graph above shows the average temperature from January 2000 to December 2012. From this plot, there is no clear pattern but there seems to be a trend of increasing temperatures.
----- End 

1.   Create a plot of the monthly average temperatures starting at January 1990 
and ending on December 1999. Do you see any pattern to the data? Are temperatures 
rising? Which months had the highest and lowest temperatures?

```{r monthly_averages, echo=FALSE}

# Average temp over 9 year period
temp_m_90s<- temp_matrix[, 110:119]

 plot(1:120, as.vector(temp_m_90s), ylab = "Temperature (F)", xlab ="Months (Since Jan 1990)", pch = 19, col = "dark blue", cex = 0.3, main = "Monthly Average Temperature")
 
```


1.   Plot the monthly average temperature of every month in the 1890s. What is 
the mean temperature of this decade? How does this decade compare to the 1990s?

```{r 1980s_plot, echo=FALSE}
temp_m_1880s<- temp_matrix[, 10:19]
plot(1:120, as.vector(temp_m_1880s), ylab = "Temperature (F)", xlab ="Months in the 1890's", pch = 19, col = "brown", cex = 0.3, main = " 1890's Monthly Average Temperature")

```

Compared to the 1890s, the 1990s average temperatures were a few degrees higher.


1.   Create plots of the monthly averages of the first 44 years, the second 44 
years, and the last 43 years. Do you see any warming trends in any of these 
plots? What is the average temperature of each of these time periods? Remember 
to use the previously defined functions!

```{r first_44, echo=FALSE}

# First 44-years 
temp_m_first_44<- temp_matrix[, 1:44]
plot(1:528, as.vector(temp_m_first_44), ylab = "Temperature (F)", xlab ="Months", pch = 19, col = "brown", cex = 0.3, main = "Monthly Average Temperature (1881-1925)") 

# Average temp for first 44 years
av_temp_first_44 <- signif(mean(as.vector(temp_m_first_44)), digits = 4)
  
# Second 44-years
temp_m_second_44<- temp_matrix[, 45:89]
plot(1:528, as.vector(temp_m_first_44), ylab = "Temperature (F)", xlab = "Months", pch = 19, col = "blue", cex = 0.3, main = "Monthly Average Temperature (1926-1970)")

# Average temp for second 44 years
av_temp_second_44 <- signif(mean(as.vector(temp_m_second_44)), digits = 4)

# Last 43 years
temp_m_last_43<- temp_matrix[, 90:132]
plot(1:528, as.vector(temp_m_first_44), ylab = "Temperature (F)", xlab ="Months", pch = 19, col = "dark red", cex = 0.3, main = "Monthly Average Temperature (1971-2012)")

# Average temp for last 43-years
av_temp_last_43<- signif(mean(as.vector(temp_m_last_43)), digits = 4)

```

The average temperature for 1881 to 1925 time period was `r av_temp_first_44`.  

The average temperature for 1926 to 1970 time period was `r av_temp_second_44`.  

The average temperature for 1971 to 2012 time period was `r av_temp_last_43`.  

Overall average temperature is increasing over each time frame.

<div id="license">
Adapted from material developed by Olaf Hall-Holt and Paul Roback for CS 125 at St. Olaf College.
</div>
