---
title: "Vectors: practice II"
author: "R Workshop"
output:
  html_document:
    css: ../workshop.css
    highlight: pygments
    theme: cerulean
  pdf_document: default
---

```{r, echo=FALSE, warning=FALSE, message=FALSE}
require(knitr)
opts_chunk$set(warning=FALSE,message=FALSE)
```


## Sorting Vectors

1.   

In addition to its use on vectors, `order()` can also be used to sort (or
I suppose order!) dataframes.  Consider one of R's preloaded datasets, from
Motor Trend Cars.

```{r, echo=FALSE}
data(mtcars)
mtcars
```

say you wanted to look through this list of cars and order them in a way in a way
that showcases which ones have the best `mpg`, then, the ones that have the 
highest `hp`.  Using `order()`, rearrange this data frame such that it's ordered
first by `mpg` (low to high) and then by `hp` (high to low).

```{r, echo=FALSE}
# Order by  ascending mpg and descending hp
new.mtcars<-mtcars[order(mtcars$mpg, -mtcars$hp),]
```
* * *

## Changing Vector Elements

Please paste the following commands into your Rstudio session, in order to set up the following exercises.

```{r}
require('ggplot2')
load(url('http://www.stolaf.edu/people/olaf/cs125/hw9.RData'))
```

Let's see whether the drawing function is working correctly. Try pasting in the following command, and see what it does:

```{r}
draw(p0x, p0y)
```

If all goes well, a simple sketch of Calvin should appear.

1.   

If you now draw the next set of vectors:

```{r}
draw(p1x, p1y)

# To fix this error, remove the y coordinates that are larger than 70 but smaller than 100
# There are 50 indexes - greater than 70 but smaller than 100
which(p1y>70 & p1y<100)

# Remove these indexes from the x and y values so they are not mapped
new.p1x<-p1x[-which(p1y>70 & p1y<100)]
new.p1y<-p1y[-which(p1y>70 & p1y<100)]

# Redraw them 
draw(new.p1x, new.p1y)
                                   
```

you will see Calvin and his mirror reflection, except that there is a problem with the data: some garbage points are mixed in with the correct data, resulting in a lot of spikes that reach reach between 75 and 100 in the y direction. Please fix this problem, so that drawing `p1x` with `p1y` results in a clean rendering of Calvin and his reflection.

1.   

Let's take a look at the second example:

```{r}
draw(p2x, p2y)

# This is Hobbes' fourth image. 

# Let's calculate the diff in y between the upper-right image and the bottom-left image and record it
diff.y2<-p2y[p2y>60 & p2y<100] - p2y[p2y<60 & p2y>0]

# The difference is consistenly 50
draw(p2x[400:800], p2y[400:800]+50)

# Since we have 3 images, let's divide the total number of x and y points by 3 and get the y-cordinates of the upper-left image and x-coordinates of the bottom-right image.
new.x<-c(p2x, p2x[400:800])
new.y<-c(p2y, p2y[400:800]+50)

draw(new.x, new.y)
# Add this to the vectors p2x and p2y and we are good to go 

```

here, we almost have four duplicates of Hobbes in a square configuration, but the upper right drawing is missing. Please duplicate one of the drawings of Hobbes, and place it in the upper right corner (by modifying the `p2x` and `p2y` vectors, that is).

1.   

Let's take a look at the third example:

```{r}
draw(p3x, p3y)

# What we want to do is to transform the first image to the second by swaping the values of x for y and vs

# Original image
draw(p3x[1315:length(p3x)], p3y[1315:length(p3x)])
# Final image 
draw(p3y[1315:length(p3x)], p3x[1315:length(p3x)])

# Reassign the values
new.p3x<-p3x
new.p3y<- p3y
new.p3x[1315:length(p3x)]<-p3y[1315:length(p3x)]
new.p3y[1315:length(p3x)]<-p3x[1315:length(p3x)]

# Redraw the picture 
draw(new.p3x, new.p3y)
```

here, Calvin is running around a circle, but the drawing closest to the origin is backwards. In particular, the x and y coordinates need to be flipped to restore symmetry. As before, please fix this problem.


<div id="license">
Adapted from material developed by Olaf Hall-Holt and Paul Roback for CS 125 at St. Olaf College.
</div>